Title
=====
Granular population estimation in settlements using passively collected data:
example of smart street sensor project.

Acknowledgement
===============
- Supervisors
- Collegues
- Family and Friends
- OpenSource projects
  (esp. GNU Linux and R)

Abstract
========
 - data is everywhere but no information
 - need to understand population in detail
 - this research connects this two quests
 - wifi probes in builtenvironment = detailed footfall information
 - we conduct experiments collect two sets of data.
 - we process the data with different methods to get footfall
 - we establish the application of the data with series of examples

Impact Statement
================
 - we live in era of explosion of data.
 - everyone is looking for ways to use data
 - this research looks in to one such data set
 - deals with collection, processing and application.
 - we develop a opensource toolkit for doing all the above
 - the research resulted in publications
 - Outputs with cdrc and subsequent use of the data 
 - methods communicated to data partner for industry application

List of Outputs and Activities
==============================
 - Amsterdam Conference
 - Mexico Conference
 - Data natives
 - Retail futures Conference 
 - CDRC footfall dashboard
 - CDRC footfall indicator
 - CDRC footfall atlas
 - IJGIS paper
 - Transfer Entropy
 - Medium data paper
 - Workshop talk - Tank
 - CDRC book
 - Humans book
 - footfall package R

Contents
========

List of Figures
===============

List of Tables
==============

Introduction
============

1) From buildings to information
------------------------------
We talk about the theory of cities and built environment.
we start from how these have been percieved as function of the
form and gradually changed to people, activity, economy and information.
Built environment is manifestation of information exchange that happens
in them. we talk about the change in theory regarding this. we talk about
how this information exchange has been becoming more and more open and
the oppotunity it provides us planners, geographers and researchers
to understand these thigns better.

2) Strutured vs unstructured data
---------------------------------
Along with the information age there is an explosion of open data.
The data collection has changed from strucutred high effort activity
to low effort scraping activity. The data generated by scraping is
unprecedented and staggering. Most of the Big-data research has gone into
this in the past decade. Disadvantages of structured data which the
unstructured data fills. This is changing how we view, undestand and experience
the world. Some of these datasets fall into this unique medium size category 
as well which are neither big data nor trivial. There is
a need for methods and tools to collect, convert and use these data.

3) Wireless and Wi-Fi
---------------------
Talk about the ubiquity of the mobile technology. Everyone has a device
which connects them with world wirelessly. Major ones cellphone and Wi-Fi
Wi-Fi is uniquely placed in between Cellphone and Bluetooth. the design
of Wi-Fi gives us amazing opprtunity. This has been done before for the past
decade by loads. The privacy advocacy has become a new thing. The change is
from both ends. Collectors are regulated, cellphones are getting sneakier.
Need for method to collect data and analyse it without compromising on privacy.
The conversion of this unstructured data into something tangible and measurable
is not a trivial problem. There are loads of such data and measurements.
examples - banking vs economic activity, oyster card data vs movement.

4) Application
--------------
The potential use of such information is immense. give examples of
smart city paradigm and connected city where realtime census is possible.
We can not only take snapshot of the state of the city, we can record and
understand the built environment as living, breathing organism. 
The insights we get by combining this information with other similar
info is more than sum of their parts. It can revolutionise understanting
, planning, policy etc, urban management and finally industry such as
retail, transportation etc. city mapper, sharing economy etc.
